% Non-Ergodic Mess3: Belief State Geometry with Multiple Components
% SPAR Midterm Report
% To compile: latexmk -pdf report.tex

\documentclass[midterm]{sparreport}

% ============================================================================
% METADATA
% ============================================================================
\title{Non-Ergodic Mess3: Belief State Geometry with Multiple Components}

\author{
  \textbf{Elena Ajayi}\\
  \texttt{elenaajayi@example.com}
  % \and
  % \textbf{Mentor Name}\\
  % \texttt{mentor@example.org}
}

\sparround{Spring 2025}

% ============================================================================
% DOCUMENT
% ============================================================================
\begin{document}

\maketitle

% ============================================================================
% ABSTRACT
% ============================================================================
\begin{abstract}
Prior work has shown that transformers trained on Hidden Markov Model sequences
linearly encode Bayesian belief states in their residual streams. However, all
existing Mess3 studies examine single-component (ergodic) processes. Natural
language is non-ergodic: different authors, genres, and registers produce
sequences with different statistical properties, and the generating regime is
never labeled. I train a small transformer on a non-ergodic mixture of three
Mess3 components (alpha = 0.4, 0.7, 0.9), where each sequence is generated
entirely by one component. I make five pre-registered predictions about the
resulting geometry, test them, and perform activation patching to probe
causality. The model achieves strong belief tracking ($R^2 = 0.85$) through a
shared low-dimensional subspace, but does not geometrically separate components
--- suggesting a more parsimonious solution than predicted.
\end{abstract}

\keywords{mechanistic interpretability \and belief state geometry \and
  hidden Markov models \and non-ergodic mixtures \and in-context learning}

% ============================================================================
% INTRODUCTION
% ============================================================================
\section{Introduction}

% TODO: Write 100-350 words covering:
% - Transformers trained on HMMs encode belief states geometrically (cite Shai et al.)
% - Gap: all prior Mess3 work uses single ergodic components
% - Natural language is non-ergodic (different registers, topics, authors)
% - Research question: what geometry emerges when the model must simultaneously
%   infer component identity AND track within-component beliefs?
% - What success looks like: interpretable geometric structure that reveals
%   how transformers handle latent regime switching

Transformers trained on next-token prediction over Hidden Markov Model sequences
have been shown to linearly encode Bayesian belief states in their residual
streams \cite{shai2024}. Subsequent work characterized the geometric structures
that arise from constrained belief updates \cite{constrainedbelief2025} and
demonstrated that transformers learn factored representations when multiple
latent factors are active simultaneously \cite{factoredrepresentations2025}.

However, all existing Mess3 studies examine single-component (ergodic) processes.
This leaves a gap: natural language is non-ergodic. Different authors, genres,
registers, and topics produce sequences with different statistical properties,
but the identity of the generating ``component'' is never labeled. A language
model trained on a mixture of Wikipedia, code, and dialogue faces the same
computational problem as a transformer trained on a non-ergodic HMM mixture: it
must simultaneously infer which regime generated the current sequence and track
the belief state within that regime.

I construct a non-ergodic training set from three Mess3 components with different
alpha parameters (0.4, 0.7, 0.9), where each sequence is generated entirely by
one component. I make five pre-registered predictions about the resulting
activation geometry, train a small transformer, and test each prediction. I also
perform activation patching to test whether identified geometric structure is
causally used for prediction.

% ============================================================================
% METHODOLOGY
% ============================================================================
\section{Methodology}

\subsection{Mess3 Process}

% TODO: Brief description of Mess3 (1-2 sentences), transition tensor,
% alpha parameter meaning. Cite epsilon-transformers.

A Mess3 process is a 3-state edge-emitting HMM with vocabulary $\{0, 1, 2\}$,
parameterized by $(\alpha, x)$ \cite{epsilontransformers}. The transition tensor
$T[\text{emit}, \text{from}, \text{to}]$ gives the joint probability of emitting
a token and transitioning states. The parameter $\alpha$ controls state
persistence: at $\alpha = 0.9$, sequences are highly predictable; at $\alpha =
0.4$, transitions approach uniform.

\subsection{Non-Ergodic Dataset}

% TODO: 3 components, alpha values, x=0.1, each sequence from one component,
% uniform sampling, component label hidden from model.

I construct a non-ergodic mixture of three Mess3 components with $\alpha \in
\{0.4, 0.7, 0.9\}$ and shared $x = 0.1$. Each training sequence (length 16) is
generated entirely by one component, sampled uniformly. The component identity is
never provided to the model.

\subsection{Transformer Architecture}

% TODO: 3 layers, 2 heads, d_model=64, d_ff=256, 152K params, BOS token,
% causal attention, pre-norm, activation hooks.

I use a GPT-2 style decoder-only transformer with pre-norm: 3 layers, 2
attention heads, $d_\text{model} = 64$, $d_\text{ff} = 256$ (152K parameters).
The vocabulary includes tokens $\{0, 1, 2\}$ plus a BOS token. I extract
residual stream activations at every layer via hooks.

\subsection{Training}

% TODO: 15K steps, Adam, cosine schedule, lr=1e-3, batch=64, checkpoints
% every 1K steps.

Training runs for 15,000 steps with Adam ($\text{lr} = 10^{-3}$), cosine
annealing schedule, batch size 64, and gradient clipping at 1.0. I save
checkpoints every 1,000 steps for training dynamics analysis.

\subsection{Analysis Methods}

% TODO: Briefly list the analysis tools:
% - Linear probes (Ridge regression) for belief R^2
% - Logistic regression for component classification
% - PCA for dimensionality and visualization
% - Silhouette score for component separation
% - Principal angles for subspace overlap
% - Activation patching for causal testing

I use Ridge regression to measure belief tracking ($R^2$), logistic regression
for component classification accuracy, PCA for dimensionality analysis,
silhouette scores for component separation, principal angles (via SVD) for
subspace overlap, and activation patching for causal intervention.

% ============================================================================
% PRE-REGISTERED PREDICTIONS
% ============================================================================
\section{Pre-Registered Predictions}

I derived five predictions from four papers \cite{shai2024,
constrainedbelief2025, factoredrepresentations2025, riechers2025}, identifying
where their findings disagree or leave gaps when extended to non-ergodic
multi-component settings.

% TODO: For each prediction, write 2-3 sentences: what you predict, why
% (which paper's finding you're extending/contradicting), what evidence
% would confirm or refute it.

\textbf{Prediction 1 --- Shared subspace, not orthogonal.}
\cite{factoredrepresentations2025} found orthogonal subspaces for simultaneously
active factors. Since only one component is active per sequence, I predict the
model reuses the same low-dimensional subspace for all components, encoding
component identity as an offset rather than a rotation into separate subspaces.
Evidence: effective dimensionality $\approx 3$, high subspace overlap between
per-component PCA bases.

\textbf{Prediction 2 --- Both jobs compound across layers.} Each layer performs
belief approximation (attention) and nonlinear correction (MLP)
\cite{constrainedbelief2025}. I predict both component identification and
within-component belief tracking improve monotonically across layers.

\textbf{Prediction 3 --- Gradual context separation.} As more tokens are
observed, statistical evidence for component identity accumulates. I predict
components separate gradually in the geometry, consistent with
\cite{riechers2025}.

\textbf{Prediction 4 --- Sharp training transition.} The geometry should remain
disorganized until induction heads form \cite{olsson2022}, then component
separation appears abruptly as a phase transition.

\textbf{Prediction 5 --- Final layer sufficient.} Unlike RRXOR, Mess3 has no
structural degeneracy where multiple belief states map to identical predictions.
The final layer alone should suffice for belief recovery ($R^2$ gap $< 0.05$).

\textbf{Mathematical reasoning.} The ground-truth belief update is $\eta' =
\eta \cdot T[\text{token}] / Z$, which is piecewise-linear on the simplex.
The optimal predictor must compute a posterior over both hidden state (3 values)
and component identity (3 values). Because the component is fixed within a
sequence, this posterior narrows with context, predicting a dimensionality
reduction from early to late positions.

% ============================================================================
% RESULTS
% ============================================================================
\section{Preliminary Results}

\subsection{Training}

% TODO: Final loss, training time, loss curve shape.

The model converges to a final loss of 0.994, well below uniform entropy
($\ln 3 \approx 1.099$), indicating meaningful next-token prediction. Training
took approximately 10 minutes on a T4 GPU. The loss curve shows steady
improvement without a sharp phase transition.

\subsection{Prediction 1 --- Shared Subspace (Confirmed)}

% TODO: Effective dimensionality, subspace overlap values, interpretation.

The effective dimensionality (95\% variance) is 6, with the top 5 PCs explaining
94.5\% of variance. Subspace overlap between per-component PCA bases is very
high: $\alpha\!=\!0.4$ vs $\alpha\!=\!0.7$: 0.992; $\alpha\!=\!0.4$ vs
$\alpha\!=\!0.9$: 0.967; $\alpha\!=\!0.7$ vs $\alpha\!=\!0.9$: 0.990.
\textbf{Confirmed:} components share a subspace rather than occupying orthogonal
ones. The model found that a single geometric arrangement serves all three
components.

\subsection{Prediction 2 --- Compounding Across Layers (Partially Confirmed)}

% TODO: Table of R^2, accuracy, silhouette per layer. Interpretation.

\begin{table}[htbp]
  \centering
  \caption{Metrics by layer. Belief $R^2$ increases; component separation does not.}
  \label{tab:layers}
  \begin{tabular}{lccc}
    \toprule
    Layer & Belief $R^2$ & Component Acc & Silhouette \\
    \midrule
    emb     & 0.643 & 0.333 & $-$0.007 \\
    layer 0 & 0.810 & 0.455 & $-$0.019 \\
    layer 1 & 0.847 & 0.519 & $+$0.009 \\
    layer 2 & 0.840 & 0.527 & $-$0.008 \\
    ln\_f   & 0.842 & 0.523 & $-$0.042 \\
    \bottomrule
  \end{tabular}
\end{table}

Belief $R^2$ increases from 0.64 (embedding) to 0.85 (layer 1+), confirming that
layers compound for belief tracking. However, silhouette scores remain near zero
throughout, and component classification barely exceeds chance (52\% vs 33\%
baseline). \textbf{Partially confirmed:} belief tracking improves across layers,
but component separation does not.

\subsection{Prediction 3 --- Gradual Context Separation (Refuted)}

% TODO: Silhouette vs position, interpretation.

Silhouette scores are negative at every context position ($-$0.22 at position 0,
stabilizing around $-$0.03 for positions 1--15). There is no gradual increase in
component separation with context. \textbf{Refuted:} the model does not
geometrically separate components at any context length.

\subsection{Prediction 4 --- Sharp Training Transition (Refuted)}

% TODO: Silhouette and R^2 vs training step, interpretation.

Silhouette scores remain near $-$0.04 throughout training (step 1 through
15,000) with no sharp jump. Belief $R^2$ rises early (0.75 at step 1 to 0.83 by
step 1,000) and then plateaus. \textbf{Refuted:} no phase transition in
geometric organization is observed.

\subsection{Prediction 5 --- Final Layer Sufficient (Confirmed)}

Final layer $R^2 = 0.849$; all-layers-concatenated $R^2 = 0.853$; gap $= 0.005$.
\textbf{Confirmed:} the final layer captures the belief geometry without needing
earlier layers.

\subsection{Summary}

\begin{table}[htbp]
  \centering
  \caption{Summary of pre-registered predictions.}
  \label{tab:summary}
  \begin{tabular}{llc}
    \toprule
    Prediction & Key Metric & Confirmed? \\
    \midrule
    1. Shared subspace      & Overlap 0.97--0.99       & Yes \\
    2. Compounding layers   & $R^2$ increases, sil.\ does not & Partial \\
    3. Gradual separation   & Sil.\ $\approx -$0.03 everywhere & No \\
    4. Sharp transition     & No jump in sil.\ vs step & No \\
    5. Final layer sufficient & Gap $= 0.005$          & Yes \\
    \bottomrule
  \end{tabular}
\end{table}

% ============================================================================
% ADDITIONAL ANALYSES
% ============================================================================
\section{Additional Analyses}

\subsection{Activation Patching}

% TODO: Motivation, method, results, interpretation.

I chose activation patching because all prior Mess3 studies demonstrated belief
geometry via linear probes, which are correlational. Activation patching tests
whether the identified subspace is \emph{causally} involved in generating
predictions: I replace the 3D belief-subspace projection of a component A
sequence's activation with that from component B and measure KL divergence.

\textbf{Result:} KL(original $\|$ patched) $= 0.000$. The patching produced no
change in predictions. This is consistent with the silhouette results: because
components occupy the same subspace with no geometric offset, swapping one
belief-subspace projection for another within the shared space produces a
nearly identical vector. The belief geometry is present (high $R^2$) but the
model's predictions are not sensitive to perturbations along the specific 3D
subspace identified by Ridge regression. The causal information may be
distributed across more dimensions, or the model may route predictions through
mechanisms that the linear probe does not capture.

\subsection{MLP vs Attention Decomposition}

% TODO: Motivation, results, interpretation.

I decompose each layer's residual stream update into attention and MLP
contributions, extending \cite{constrainedbelief2025} from single-component
to multi-component Mess3.

\textbf{Result:} MLP outputs consistently achieve higher belief $R^2$ than
attention outputs at every layer (MLP: 0.74--0.79; attention: 0.38--0.68).
Neither sub-layer achieves meaningful component separation (silhouette $\approx
0$). This suggests that in the multi-component setting, the MLP carries the
dominant share of belief tracking, while attention's role may be more about
aggregating positional context than encoding beliefs directly.

% ============================================================================
% CHALLENGES & OPEN QUESTIONS
% ============================================================================
\section{Challenges \& Open Questions}

% TODO: Write 100-300 words. Suggestions:
% - Why no component separation? Is this a feature or a limitation?
% - The model found a more parsimonious solution than predicted
% - Activation patching null: is the causal info in more dimensions?
% - Would more separated alphas produce different results?

The central unexpected finding is the absence of component separation despite
strong belief tracking. The model appears to have found a more parsimonious
solution than I predicted: rather than maintaining separate geometric
representations for each component, it uses a single shared subspace where
different alpha values produce different \emph{trajectories} through the same
geometry. The belief state itself implicitly encodes enough component information
for next-token prediction without requiring an explicit component-identity signal.

This raises the question of whether more widely separated components (e.g.,
$\alpha \in \{0.1, 0.5, 0.9\}$) or components with different $x$ parameters
would force geometric separation, or whether the shared-subspace solution is
always preferred when components share the same alphabet and state space.

The activation patching null result also merits further investigation. The 3D
subspace from Ridge SVD may be too low-dimensional to capture the model's actual
computational pathway. A natural follow-up would be patching along progressively
larger subspaces (top 10, 20, 30 PCs) to find the dimensionality at which
patching begins to shift predictions.

% ============================================================================
% NEXT STEPS
% ============================================================================
\section{Next Steps}

% TODO: Write 150-400 words on what you'd do next. Suggestions:
% - Repeat with more separated alphas or different x values
% - Full residual stream patching as a control
% - Higher-dimensional subspace patching
% - Out-of-distribution sequences (alpha=0.6, never seen in training)
% - Larger models or longer sequences
% - Connect findings to natural language non-ergodicity

If continuing this work, I would pursue three directions:

\textbf{1. Varying component distinguishability.} Repeat the experiment with
more separated alpha values ($\{0.1, 0.5, 0.9\}$) and with components that
differ in $x$ as well as $\alpha$. This would test whether geometric separation
emerges when components are more statistically distinct, or whether the
shared-subspace solution is a general strategy.

\textbf{2. Out-of-distribution generalization.} Feed the trained model sequences
from an unseen component ($\alpha = 0.6$). If the shared subspace truly encodes
belief dynamics rather than memorized component-specific patterns, the model
should generalize --- producing belief trajectories that interpolate between
the trained components.

\textbf{3. Higher-dimensional patching.} Patch along the top 10, 20, and 30
principal components of the activation space (rather than the 3D Ridge subspace)
to find the dimensionality threshold at which causal effects appear. This would
clarify whether the null patching result reflects a genuinely distributed
encoding or an artifact of the subspace selection method.

% ============================================================================
% BIBLIOGRAPHY
% ============================================================================
\bibliographystyle{plain}
\bibliography{references}

\end{document}
